Measuring Directionality in Double Beta Decay and Neutrino Interactions with Kiloton-Scale Scintillation Detectors

### 2013-07-18 Direction.tex
Added Reference for speed of light in glass and added sentence below the corresponding equation. --Christoph

A good liquid scintillator will produce $\sim$10,000 photons
isotropically per MeV of deposited energy. Although less abundant,
Cerenkov light will be produced as well if a particle is moving faster
than the speed of light in the medium.  This light is emitted in a
cone pointed in the direction of the particles trajectory and with a
continuous spectrum weighted toward shorter wavelengths
but extending well into the red. The spectrum is described by:
\begin{equation}
\label{eqCerenkov}
\frac{dN}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} ( 1 - \frac{1}{\beta^2 n(\lambda)^2} )
\end{equation}
where $n(\lambda)$ is the wavelength-dependent index of refraction and
$\beta$ is the velocity of the incoming particle. The Cerenkov light
produced at wavelengths shorter than the absorption cutoff of the
scintillator will be absorbed and re-emitted as isotropic light, but
wavelengths longer than this cutoff will propagate across the
detector, retaining their directional information. The yield is
roughly 60 photons per MeV, assuming a 400~nm absorption
cutoff\cite{qdot}. These undisturbed Cerenkov photons will have timing
determined by the group velocity in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) + dn(\lambda)/dlog(\lambda)}.
\end{equation}
The longer wavelengths Cerenkov photons typically arrive before
the scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelength, it should be
possible to separate the directional Cerenkov light and the isotropic
scintillation light in time and then reconstruct the direction of the
initial particle.

@@

%deprecated text: fake vertex reconstruction not needed anymore. 
%To this end, we use the KamLAND vertex resolution of 12.0~cm \cite{tbd} ($\sigma$ in one dimension) to randomly draw a vertex around the known true vertex for each event. This vertex is subsequently used as a reconstructed vertex and consequently, a time of flight (TOF) correction relative to the center of the sphere is applied for each photon hit depending on the distance between the reconstructed vertex and the hit position on the sphere. For the TOF correction a single effective speed of light is used. This effective speed was extracted at the mean wavelength of both Scintillation and Cerenkov photons (408~nm). The 12.0~cm vertex resolution corresponds to a time resolution broadening of about 0.6~ns ($\sigma$). Note that the vertex resolution itself is dependent on scintillator and detector properties. For example, the PMTs used in KamLAND have a TTS of 1.28~ns ($\sigma$) and significant improvement of the vertex resolution could be possible with faster photodetectors. In section \ref{reconstruction_sec}, we discuss first results on more realistic reconstruction.
Before we discuss the simulation results for different simulation settings in the following sections, we highlight the effects which contribute to the timing of the scintillator detector system. First, the simulated travel time of the initial 5~MeV electron is between 0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the scintillation light emission follows a scintillator-specific distribution characterized by rise and decay time(s). Before the solutes in liquid scintillator can emit optical photons, the energy has to be transfered from the solvent to the solute. The time constant of this energy transfer accounts for a rise time in scintillation light emission. Past neutrino experiments were not highly interested in the effect of scintillation rise time which is the reason why there is a lack of accurate numbers. We assume a rise time of 1~ns, more detailed studies are needed in the future. The two time constants used to describe the falling edge of the scintillator emission time distribution (quoted above) are values specific to the KamLAND scintillator. Third, chromatic dispersion turns out to be an important effect in a 6.5~m diameter detector at the level of precision needed for direction reconstruction. Due to the wavelength-dependence of the refractive index the speed of light in the scintillator (the group velocity is used in Geant4 for normal dispersion) is also wavelength dependent. In order to study the size of this effect, we extracted results from a simplified simulation of 5~MeV electrons at the center of the sphere where we used instantaneous scintillation and did apply neither TTS nor TOF correction. The true time distributions of photoelectrons were analyzed for scintillation light and Cerenkov light separately. Photoelectrons coming from Cerenkov light are created about 0.5~ns earlier than PEs from scintillation light on average. The RMS values for both the Cerenkov and scintillation light true PE time distributions are both about 0.5~ns. Note that these numbers include the effect of the finite electron travel time.

@@

In Sections \ref{detector_timing_sec} to \ref{edep_size_sec}, we comparatively study the photoelectron timing for different detector properties. We follow the idea to increase discrimination between Cerenkov and scintillation light by detector timing. The main quantities provided by the Geant4 simulation are the photoelectron hit positions and the detection times after TTS has been applied. In section \ref{reconstruction_sec} these quantities are used for event reconstruction.
%%statement that rise time is not super-crucial (see April 11 talk)
%% include wavelength spectra, emission spectra, absorption (?), refractive index  ?
%% Other settings: physics list, theCerenkovProcess->SetMaxNumPhotonsPerStep(20);theCerenkovProcess->SetMaxBetaChangePerStep(10.0); theCerenkovProcess->SetTrackSecondariesFirst(true); theScintillationProcess->SetScintillationYieldFactor(1.); theScintillationProcess->SetTrackSecondariesFirst(true); theScintillationProcess->SetFiniteRiseTime(true); particle gun, 5MeV e- in x direction from center, random polarization
%% discussion: can chromatic dispersion be disentangled? With very high statistics maybe yes (see Ben Monreal, Matt Wetstein).
%% discussion: many scintillator properties depend on each other. Optimization of scintillator properties could lead to better results for directionality reconstruction. i.e. slower scintillation (but vertex reco worse!, LY tends to be lower) , lower wavelength, 
%% idea: filter for part of the PMTs or two kinds of PMTs: separation by wavelength
%% coverage is just a reduction by factor 0.22 (flat over wavelengths etc.), not studied in the paper?  
%% at the 0.1 ns level in a 6.5 m detector one has to be careful with the input of nref. If the slope is not smooth enough the group velocity calcuation is creating wavelength-dependent distortions. 
%% Reflections and absorption/reemission and scattering will shift events to higher times were they will likely not pass the time cut. They will effectively change the scintillation/Cerenkov ratio. Scattering might distort the angular variables and is subject of future work. 
%% In small det. the chromatic dispersion doesn't help too much. (In the small detector the detector properties are the driving factor.)
%% discussion: some scintillator and detector properties have complicated effects on the directionality, also other requirements have to be considered (high LY, high resolutionc). Multiparameter optimization problem.  
%% nuebar scattering --> one of next papers

\section{Detector Timing}
\label{detector_timing_sec}
\begin{figure*}
        \begin{center}
        \subfigure[Default simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[Increased TTS (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[Red-sensitive photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{PE times after TTS application for the simulation of 1000 electrons (5~MeV) with different settings. PEs from Cerenkov light (black) and scintillation light (grey) are compared. The dashed vertical line illustrates a time cut at 34~ns. (a) Default simulation including bialkali photocathode and TTS = 0.1~ns ($\sigma$). The number of PEs per event after the 34.0~ns time cut is 171 from scintillation and 108 from Cerenkov. (b) Default simulation settings but for TTS = 1.28~ns (KamLAND 17 in. PMTs). The number of PEs per event after the 34.0~ns time cut is 349 from scintillation and 88 from Cerenkov. (c) Default simulation settings but for GaAsP photocathode. The number of PEs per event after the 34.0~ns time cut is 50.8 from scintillation and 171 from Cerenkov. \label{time_plots_comparison}}
        \end{center}
\end{figure*}
%%\begin{figure}
%%        \begin{center}
%%        \includegraphics[scale=0.40]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}
%%        \caption[]{PE times after TTS application for the simulation of 1000 electrons (5~MeV) with default settings: Detector diameter = 6.5~m, bialkali photocathode, KamLAND scintillator emission spectrum, TTS = 0.1~ns ($\sigma$). PEs from Cerenkov light (black) and scintillation light (red) are compared. The number of PEs per event after a 34.0~ns time cut is 171 from scintillation and 108 from Cerenkov. \label{6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME}}
%%        \end{center}
%%\end{figure}
\begin{figure}
        \begin{center}
        \includegraphics[scale=0.40]{graphs/cos_psi_comp_h.pdf}
        \caption[]{Angular distribution of photoelectron hits relative to the original electron direction for 1000 events (5~MeV electrons). Default simulation settings were used. A 33.0~ns time cut (black) is compared with a 34.0~ns time cut (red). \label{cerenkov_cone}}
        \end{center}
\end{figure}
%%\begin{figure}
%%        \begin{center}
%%        \includegraphics[scale=0.40]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}
%%        \caption[]{PE times after TTS application for the simulation of 1000 electrons (5~MeV) with TTS = 1.28~ns: Detector diameter = 6.5~m, bialkali photocathode, KamLAND scintillator emission spectrum, TTS = 1.28~ns ($\sigma$). PEs from Cerenkov light (black) and scintillation light (red) are compared. The number of PEs per event after a 34.0~ns time cut is 349 from scintillation and 88 from Cerenkov. \label{6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME}}
%%        \end{center}
%%\end{figure}

First we discuss the default simulation settings which are described in the previous section. Figure \ref{time_plot_comparison} (a) shows the TTS-smeared photoelectron detection time for 1000 simulated electrons with 5~MeV energy in the center of the detector with initial momentum direction coinciding with the x-axis. The photoelectrons induced by Cerenkov light arrive earlier, as expected due to the instantaneous emission and the differences in the photon speed. There is however overlap with the PE time distribution coming from scintillation light. In order to compare simulations with different parameters to each other a fixed time cut $t \leq$ 34.0~ns (or 33.0~ns?) is applied (using the truth information of the electron starting time) instead of an event-by-event time reconstruction (see Section \ref{reconstruction_sec} for reconstruction results). For the default simulation case, the number of PEs per event coming from Cerenkov light after the 34.0~ns time cut (108) is 98~\% of the total number of PEs from Cerenkov light (110). For scintillation the number of PEs after the time cut (171) is only 3.2~\% of the total scintillation-induced PEs (5445).

>>>>

A good liquid scintillator will produce $\sim$10,000 photons
isotropically per MeV of deposited energy. Although less abundant,
Cerenkov light will be produced as well if a particle is moving faster
than the speed of light in the medium.  This light is emitted in a
cone pointed in the direction of the particles trajectory and with a
continuous spectrum weighted toward shorter wavelengths
but extending well into the red. The spectrum is described by:
\begin{equation}
\label{eqCerenkov}
\frac{dN}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} ( 1 - \frac{1}{\beta^2 n(\lambda)^2} )
\end{equation}
where $n(\lambda)$ is the wavelength-dependent index of refraction and
$\beta$ is the velocity of the incoming particle. The Cerenkov light
produced at wavelengths shorter than the absorption cutoff of the
scintillator will be absorbed and re-emitted as isotropic light, but
wavelengths longer than this cutoff will propagate across the
detector, retaining their directional information. The yield is
roughly 60 photons per MeV, assuming a 400~nm absorption
cutoff\cite{qdot}. These undisturbed Cerenkov photons will have timing
determined by the group velocity \cite{group_velocity_article} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) + dn(\lambda)/dlog(\lambda)}.
\end{equation}
In Geant4 optical photons are assigned the group velocity in the wavelength 
region of normal dispersion ($dn(\lambda)/dlog(\lambda)$>0) which is relevant to 
our study. 
The longer wavelengths Cerenkov photons typically arrive before
the scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelength, it should be
possible to separate the directional Cerenkov light and the isotropic
scintillation light in time and then reconstruct the direction of the
initial particle.

@@

%deprecated text: fake vertex reconstruction not needed anymore. 
%To this end, we use the KamLAND vertex resolution of 12.0~cm \cite{tbd} ($\sigma$ in one dimension) to randomly draw a vertex around the known true vertex for each event. This vertex is subsequently used as a reconstructed vertex and consequently, a time of flight (TOF) correction relative to the center of the sphere is applied for each photon hit depending on the distance between the reconstructed vertex and the hit position on the sphere. For the TOF correction a single effective speed of light is used. This effective speed was extracted at the mean wavelength of both Scintillation and Cerenkov photons (408~nm). The 12.0~cm vertex resolution corresponds to a time resolution broadening of about 0.6~ns ($\sigma$). Note that the vertex resolution itself is dependent on scintillator and detector properties. For example, the PMTs used in KamLAND have a TTS of 1.28~ns ($\sigma$) and significant improvement of the vertex resolution could be possible with faster photodetectors. In section \ref{reconstruction_sec}, we discuss first results on more realistic reconstruction.
\begin{figure*}[tbh]
        \begin{center}
        \subfigure[Default simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[Increased TTS (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[Red-sensitive photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{PE times after TTS application for the simulation of 1000 electrons (5~MeV) with different settings. PEs from Cerenkov light (black) and scintillation light (grey) are compared. The dashed vertical line illustrates a time cut at 34~ns. (a) Default simulation including bialkali photocathode and TTS = 0.1~ns ($\sigma$). The number of PEs per event after the 34.0~ns time cut is 171 from scintillation and 108 from Cerenkov. (b) Default simulation settings but for TTS = 1.28~ns (KamLAND 17 in. PMTs). The number of PEs per event after the 34.0~ns time cut is 349 from scintillation and 88 from Cerenkov. (c) Default simulation settings but for GaAsP photocathode. The number of PEs per event after the 34.0~ns time cut is 50.8 from scintillation and 171 from Cerenkov. \label{time_plots_comparison}}
        \end{center}
\end{figure*}
Before we discuss the simulation results for different simulation settings in the following sections, we highlight the effects which contribute to the timing of the scintillator detector system. First, the simulated travel time of the initial 5~MeV electron is between 0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the scintillation light emission follows a scintillator-specific distribution characterized by rise and decay time(s). Before the solutes in liquid scintillator can emit optical photons, the energy has to be transfered from the solvent to the solute. The time constant of this energy transfer accounts for a rise time in scintillation light emission. Past neutrino experiments were not highly interested in the effect of scintillation rise time which is the reason why there is a lack of accurate numbers. We assume a rise time of 1~ns, more detailed studies are needed in the future. The two time constants used to describe the falling edge of the scintillator emission time distribution (quoted above) are values specific to the KamLAND scintillator. Third, chromatic dispersion turns out to be an important effect in a 6.5~m diameter detector at the level of precision needed for direction reconstruction. Due to the wavelength-dependence of the refractive index the speed of light in the scintillator (the group velocity is used in Geant4 for normal dispersion) is also wavelength dependent. In order to study the size of this effect, we extracted results from a simplified simulation of 5~MeV electrons at the center of the sphere where we used instantaneous scintillation and did apply neither TTS nor TOF correction. The true time distributions of photoelectrons were analyzed for scintillation light and Cerenkov light separately. Photoelectrons coming from Cerenkov light are created about 0.5~ns earlier than PEs from scintillation light on average. The RMS values for both the Cerenkov and scintillation light true PE time distributions are both about 0.5~ns. Note that these numbers include the effect of the finite electron travel time.

@@

In Sections \ref{detector_timing_sec} to \ref{edep_size_sec}, we comparatively study the photoelectron timing for different detector properties. We follow the idea to increase discrimination between Cerenkov and scintillation light by detector timing. The main quantities provided by the Geant4 simulation are the photoelectron hit positions and the detection times after TTS has been applied. In section \ref{reconstruction_sec} these quantities are used for event reconstruction.
%%statement that rise time is not super-crucial (see April 11 talk)
%% include wavelength spectra, emission spectra, absorption (?), refractive index  ?
%% Other settings: physics list, theCerenkovProcess->SetMaxNumPhotonsPerStep(20);theCerenkovProcess->SetMaxBetaChangePerStep(10.0); theCerenkovProcess->SetTrackSecondariesFirst(true); theScintillationProcess->SetScintillationYieldFactor(1.); theScintillationProcess->SetTrackSecondariesFirst(true); theScintillationProcess->SetFiniteRiseTime(true); particle gun, 5MeV e- in x direction from center, random polarization
%% discussion: can chromatic dispersion be disentangled? With very high statistics maybe yes (see Ben Monreal, Matt Wetstein).
%% discussion: many scintillator properties depend on each other. Optimization of scintillator properties could lead to better results for directionality reconstruction. i.e. slower scintillation (but vertex reco worse!, LY tends to be lower) , lower wavelength, 
%% idea: filter for part of the PMTs or two kinds of PMTs: separation by wavelength
%% coverage is just a reduction by factor 0.22 (flat over wavelengths etc.), not studied in the paper?  
%% at the 0.1 ns level in a 6.5 m detector one has to be careful with the input of nref. If the slope is not smooth enough the group velocity calcuation is creating wavelength-dependent distortions. 
%% Reflections and absorption/reemission and scattering will shift events to higher times were they will likely not pass the time cut. They will effectively change the scintillation/Cerenkov ratio. Scattering might distort the angular variables and is subject of future work. 
%% In small det. the chromatic dispersion doesn't help too much. (In the small detector the detector properties are the driving factor.)
%% discussion: some scintillator and detector properties have complicated effects on the directionality, also other requirements have to be considered (high LY, high resolutionc). Multiparameter optimization problem.  
%% nuebar scattering --> one of next paper

\section{Detector Timing}
\label{detector_timing_sec}
%%\begin{figure}
%%        \begin{center}
%%        \includegraphics[scale=0.40]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}
%%        \caption[]{PE times after TTS application for the simulation of 1000 electrons (5~MeV) with default settings: Detector diameter = 6.5~m, bialkali photocathode, KamLAND scintillator emission spectrum, TTS = 0.1~ns ($\sigma$). PEs from Cerenkov light (black) and scintillation light (red) are compared. The number of PEs per event after a 34.0~ns time cut is 171 from scintillation and 108 from Cerenkov. \label{6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME}}
%%        \end{center}
%%\end{figure}
%%\begin{figure}
%%        \begin{center}
%%        \includegraphics[scale=0.40]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}
%%        \caption[]{PE times after TTS application for the simulation of 1000 electrons (5~MeV) with TTS = 1.28~ns: Detector diameter = 6.5~m, bialkali photocathode, KamLAND scintillator emission spectrum, TTS = 1.28~ns ($\sigma$). PEs from Cerenkov light (black) and scintillation light (red) are compared. The number of PEs per event after a 34.0~ns time cut is 349 from scintillation and 88 from Cerenkov. \label{6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME}}
%%        \end{center}
%%\end{figure}

First we discuss the default simulation settings which are described in the previous section. Figure \ref{time_plot_comparison} (a) shows the TTS-smeared photoelectron detection time for 1000 simulated electrons with 5~MeV energy in the center of the detector with initial momentum direction coinciding with the x-axis. The photoelectrons induced by Cerenkov light arrive earlier, as expected due to the instantaneous emission and the differences in the photon speed. There is however overlap with the PE time distribution coming from scintillation light. In order to compare simulations with different parameters to each other a fixed time cut $t \leq$ 34.0~ns (or 33.0~ns?) is applied (using the truth information of the electron starting time) instead of an event-by-event time reconstruction (see Section \ref{reconstruction_sec} for reconstruction results). For the default simulation case, the number of PEs per event coming from Cerenkov light after the 34.0~ns time cut (108) is 98~\% of the total number of PEs from Cerenkov light (110). For scintillation the number of PEs after the time cut (171) is only 3.2~\% of the total scintillation-induced PEs (5445).  
\begin{figure}[bt]
        \begin{center}
        \includegraphics[scale=0.40]{graphs/cos_psi_comp_h.pdf}
        \caption[]{Angular distribution of photoelectron hits relative to the original electron direction for 1000 events (5~MeV electrons). Default simulation settings were used. A 33.0~ns time cut (black) is compared with a 34.0~ns time cut (red). \label{cerenkov_cone}}
        \end{center}
\end{figure}
